Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-10 optim.tex
edit application #2

Here we explore the potential for a jellyfish optimisation to be
integrated with the guided tour. Section~\ref{sec-background} explains
the optimisation that is used in the current the projection pursuit
guided tour. Section~\ref{sec-theory} provides more details on the
jellyfish optimiser and formalises several characteristics of projection
pursuit indexes that are help to measure optimiser performance.
Section~\ref{sec-simulation} describes a simulation study on performance
of the jellyfish for several types of data and index functions.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, opacitybacktitle=0.6, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colback=white, bottomtitle=1mm, breakable, toprule=.15mm, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toptitle=1mm, coltitle=black, leftrule=.75mm, left=2mm, arc=.35mm]

@@

\section{Application {[}Di and Sherry{]}}\label{sec-simulation}

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search better optimiser and explores how it behaves
with different combinations of hyper-parameter. The second example
studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with JS hyper-parameters, on the outcome
of the optimisation.

\subsection{Performance of JS optimiser in pipe-finding
problems}\label{performance-of-js-optimiser-in-pipe-finding-problems}

@@

\caption{\label{fig-proj}Projections found by the jellyfish and search
better optimisers at each 10th quantile across 50 simulations. The
projection pursuit problem is to find the pipe shape using the holes
index in the 6, 8, 10, and 12-dimensional spaces. The JS optimiser uses
100 jellyfishes and a maximum number of tries of 100. The search better
optimiser uses a maximum of 1000 tries in each step of random sampling
step before the algorithm terminates. In the 6-D data space, the JS
optimiser always find a clear pipe shape while the search better
optimiser also find the pipe shape but with a wide rim. At higher data
dimension, the JS optimiser always find a higher index value and a
clearer pipe shape across all the quantiles.}

@@

\subsection{Connecting optimisation success with its properties and
jellyfish
parameters}\label{connecting-optimisation-success-with-its-properties-and-jellyfish-parameters}

The optimisation properties (smoothness, squintability, and speed)
proposed in the Section~\ref{sec-theory} provide numerical measures to
characterise the difficulty of the optimisation for a given projection
pursuit problem. This example investigates how these measures, combined
with the jellyfish parameters, affect the success of the jellyfish
optimisers and how they can guide the decisions made in hyperparameter
tuning with the jellyfish optimiser. Simulations are conducted with
different jellyfish optimisers to obtain the proportion of successful
optimisation across a collection of scenarios, for which smoothness,
squintability, and speed are calculated. A generalised linear model is
used to link the proportion of success rate with the jellyfish
parameters and the optimisation properties.

In addition to the pipe-finding problem, also investigated is the
sine-wave finding problem in 6D and 8D spaces with six indices
considered: \texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC},
\texttt{TIC}, \texttt{spline}, and \texttt{stringy}. Combining with
different jellyfish optimiser parameters (\texttt{n\_jellies} and
\texttt{max\_tries}), a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, the
jellyfish optimiser is run fifty (50) times and summary statistics are
calculated on the best index value found across all 50 runs and the
proportion of runs that find a close (with a difference less than 0.05)
best index value (\texttt{P\_J\_hat}).

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance to the
optimal basis are calculated for each basis to fit a Gaussian process
model. For squintability, fifty random bases are sampled and
interpolated to the optimal basis with a step size of 0.005. The
interpolated bases are first binned with a bin width of 0.005 to average
the index values before feeding into a 4-parameter scaled logistic
function, estimated by non-linear least square, to obtain the
squintability measure.

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated for each case from the Gaussian process (variance, range,
smooth, and nugget) for smoothness and the scaled logistic function
(theta1 to theta4, and squint) for squintability. The column ``smooth''
and ``squint'' are used as the smoothness and squintability measure. The
low squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure. {[}TODO: the stringy squintability doens't make sense{]}

Table~\ref{tbl-mod-data} presents the processed data for modelling after
augmenting the jellyfish parameters (\texttt{n\_jellies} and
\texttt{max\_tries}) and the average time spent on each run
(\texttt{time}) for each case. A generalised linear model is fitted with
a binomial family and a logit link function to the data to investigate
the relationship between the proportion of jellyfish success and
jellyfish parameters and the optimisation properties. Pre-processing of
the data include 1) scale the jellyfish parameters (\texttt{n\_jellies}
and \texttt{max\_tries}) by 10 for interpretation, and 2) create a
binary variable \texttt{long\_time} to flag the cases with average time
spent on each run greater than 30 seconds. Table~\ref{tbl-mod-output}
presents the model output with the estimated coefficients. The model
suggests a positive relationship between the proportion of jellyfish
success and the jellyfish parameters, \texttt{n\_jellies} and
\texttt{max\_tries}, optimisation property parameter,
\texttt{smoothness} and \texttt{squintability}, and a negative
relationship with the average time spent on each run
(\texttt{long\_time}) and the data dimension (\texttt{d}). The variable
\texttt{squintability} and \texttt{d} are significant, suggesting their
importance over jellyfish parameters to the success of the optimisation.

@@

\begin{longtable}[t]{|>{}ll>{}r|rrr>{}r|rrrr>{}r|}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (theta1 to theta4) for the pipe-finding and
sine-wave finding problems. The squint column is calculated as
\(\frac{\theta1 \theta2 \theta3}{4}\), as described in
Section~\ref{sec-squintability}. The ``smooth'' and ``squint'' column
represent the smoothness and squintability measures.}

@@

\toprule
shape & index & d & variance & range & smooth & nugget & theta1 & theta2 & theta3 & theta4 & squint\\
\midrule
pipe & holes & 6 & 0.002 & 0.408 & 2.364 & 0.212 & 1.001 & 0.860 & 3.368 & 0.823 & 0.725\\
pipe & holes & 8 & 0.000 & 0.259 & 2.373 & 0.613 & 1.001 & 0.869 & 3.264 & 0.811 & 0.710\\
pipe & holes & 10 & 0.000 & 0.144 & 2.317 & 1.831 & 1.000 & 0.885 & 3.151 & 0.806 & 0.697\\
pipe & holes & 12 & 0.000 & 0.254 & 2.173 & 0.879 & 1.000 & 0.878 & 3.345 & 0.806 & 0.734\\
sine & MIC & 6 & 0.016 & 0.100 & 2.394 & 0.087 & 0.894 & 0.571 & 1.623 & -0.024 & 0.207\\
\addlinespace
sine & MIC & 8 & 0.016 & 0.100 & 2.394 & 0.087 & 0.932 & 0.328 & 1.314 & -0.030 & 0.100\\
sine & TIC & 6 & 0.124 & 0.104 & 2.471 & 0.086 & 0.951 & 0.536 & 1.719 & -0.025 & 0.219\\
sine & TIC & 8 & 0.124 & 0.104 & 2.471 & 0.086 & 0.945 & 0.564 & 1.723 & -0.027 & 0.230\\
sine & dcor2d\_2 & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.679\\
sine & loess2d & 6 & 0.083 & 0.307 & 2.194 & 0.292 & 1.016 & 1.039 & 2.648 & 0.080 & 0.699\\
\addlinespace
sine & splines2d & 6 & 0.040 & 0.189 & 2.606 & 0.104 & 1.014 & 1.051 & 2.730 & -0.009 & 0.727\\
sine & stringy & 6 & 0.000 & 1173.035 & 1.031 & 17608.047 & 1.011 & 0.011 & 254.734 & 0.727 & 0.711\\
\bottomrule

@@

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0811}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0405}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1486}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1892}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1351}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1351}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1622}}@{}}

\caption{\label{tbl-mod-data}The first few rows of the datasets
processed for modelling. The smoothness and squintability variable are
uniquely characterised by the index function used and the data
dimension, and thus do not vary across \texttt{n\_jellies} and
\texttt{max\_tries}. The variable \texttt{P\_J\_hat}, and time are
calculated at each observation.}

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
index
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
d
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
smoothness
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
squintability
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_jellies
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
max\_tries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P\_J\_hat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
time
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
MIC & 6 & 2.394 & 0.207 & 20 & 50 & 0.12 & 2.479 secs \\
MIC & 6 & 2.394 & 0.207 & 20 & 100 & 0.24 & 8.950 secs \\
MIC & 6 & 2.394 & 0.207 & 50 & 50 & 0.52 & 5.651 secs \\
MIC & 6 & 2.394 & 0.207 & 50 & 100 & 0.64 & 13.223 secs \\
MIC & 6 & 2.394 & 0.207 & 100 & 50 & 0.76 & 19.453 secs \\
MIC & 8 & 2.394 & 0.100 & 20 & 50 & 0.08 & 2.566 secs \\
MIC & 8 & 2.394 & 0.100 & 20 & 100 & 0.08 & 4.960 secs \\

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success (\texttt{P\_J\_hat}) on optimisation properties
(\texttt{smoothness}, \texttt{squintability}, and \texttt{d}) and
jellyfish parameters (\texttt{long\_time}, \texttt{n\_jellies}, and
\texttt{max\_tries}) from the generalised linear model with a binomial
family and a logit link function. The variable \texttt{long\_time} is
derived to flag the cases with average time spent on each run greater
than 30 seconds and the variable \texttt{n\_jellies} and
\texttt{max\_tries} are scaled by 10 for interpretation.}

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
intercept & -4.672 & 4.512 & -1.036 & 0.300\\
smoothness & 1.604 & 1.530 & 1.048 & 0.294\\
squintability & 7.062 & 2.210 & 3.195 & 0.001\\
d & -0.595 & 0.260 & -2.289 & 0.022\\
long\_time & -0.851 & 1.348 & -0.632 & 0.528\\
\addlinespace
n\_jellies & 0.230 & 0.131 & 1.754 & 0.080\\
max\_tries & 0.107 & 0.153 & 0.700 & 0.484\\
\bottomrule

>>>>

Here we explore the potential for a jellyfish optimisation to be
integrated with the guided tour. Section~\ref{sec-background} explains
the optimisation that is used in the current the projection pursuit
guided tour. Section~\ref{sec-theory} provides more details on the
jellyfish optimiser and formalises several characteristics of projection
pursuit indexes that are help to measure optimiser performance.
\textbf{?@sec-simulation} describes a simulation study on performance of
the jellyfish for several types of data and index functions.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, coltitle=black, rightrule=.15mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, left=2mm, breakable, colbacktitle=quarto-callout-note-color!10!white, colback=white, toptitle=1mm, opacityback=0, opacitybacktitle=0.6, arc=.35mm, bottomtitle=1mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame]

@@

\section{Application {[}Di and Sherry{]}}\label{sec-app}

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search better optimiser and explores how it behaves
with different combinations of hyper-parameter. The second example
studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with jellyfish hyper-parameters, on the
outcome of the optimisation.

\subsection{Performance of JS optimiser in pipe-finding
problems}\label{sec-app-1}

@@

\caption{\label{fig-proj}Projections found by the jellyfish and search
better optimisers at each 10th quantile across 50 simulations. The
projection pursuit problem is to find the pipe shape using the holes
index in the 6, 8, 10, and 12-dimensional spaces. The JS optimiser uses
100 jellyfishes and a maximum number of tries of 100. The search better
optimiser uses a maximum of 1000 tries in each step of random sampling
step before the algorithm terminates. In the 6-D data space, the JS
optimiser always finds a clear pipe shape while the search better
optimiser also finds the pipe shape but with a wide rim. At higher data
dimensions, the JS optimiser finds a higher index value and a clearer
pipe shape across all the quantiles than the search better optimiser.}

@@

\subsection{Factors affecting JS optimiser success rate: optimisation
properties and jellyfish hyper-parameters}\label{sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, alongside jellyfish hyper-parameters (the number of
jellyfish and the maximum number of tries), affect the success of the JS
optimiser. Simulations are conducted to obtain the success rate across
various projection pursuit problems, characterised by shape-to-find,
data dimension, and index function used, as well as different
hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
Section~\ref{sec-theory}. A generalised linear model is used to
construct the relationship between the success rate, jellyfish
hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, the JS
optimiser is run fifty (50) times and the success rate is calculated as
in Section~\ref{sec-app-1}.

\emph{Note: should we print out the 52 cases investigated somewhere, say
appendex, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes {[}MIC \& TIC{]} x 6 configurations = 12
cases for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.}

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance (to the
optimal basis) are calculated for each random basis before fitting them
into a Gaussian process model. For squintability, fifty random bases are
sampled and interpolated to the optimal basis with a step size of 0.005.
For these interpolated bases, index values and projection distances to
the optimal basis are calculated and averaged with a bin width of 0.005.
The squintability measure is then calculated from fitting a 4-parameter
scaled logistic function of index value against projection distance,
estimated by non-linear least square.

\emph{Note: Shall I repeat the scaled logistic function with our
parameterisation here? It can be handy to interpret theta1 - theta4 but
they should've already been explained in the squintability section.}

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as
\(\frac{\theta_1 \theta_2 \theta_3}{4}\) as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta_2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure.

\emph{Note: the stringy squintability doesn't make sense}

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three pre-processing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of the JS
optimiser is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to jellyfish hyper-parameters in the optimisation sucess.

@@

\begin{longtable}[t]{|>{}ll>{}l|rrr>{}r|rrrr>{}r|>{}r|}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The ``smooth'' column and
``squint'' column, calculated as
\(\frac{\theta_1 \theta_2 \theta_3}{4}\), represent the smoothness and
squintability measures.}

@@

\toprule
 & shape & index & d & variance & range & smooth & nugget & theta1 & theta2 & theta3 & theta4 & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.408 & 2.364 & 0.212 & 0.991 & 0.860 & 3.368 & 0.172 & 0.718\\
2 & pipe & holes & 8 & 0.000 & 0.259 & 2.373 & 0.613 & 1.011 & 0.869 & 3.264 & 0.085 & 0.716\\
3 & pipe & holes & 10 & 0.000 & 0.144 & 2.317 & 1.831 & 1.011 & 0.885 & 3.151 & 0.153 & 0.705\\
4 & pipe & holes & 12 & 0.000 & 0.254 & 2.173 & 0.879 & 0.999 & 0.878 & 3.345 & 0.196 & 0.733\\
5 & sine & MIC & 6 & 0.016 & 0.100 & 2.394 & 0.087 & 0.894 & 0.571 & 1.623 & -0.024 & 0.207\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.100 & 2.394 & 0.087 & 0.932 & 0.328 & 1.314 & -0.030 & 0.100\\
7 & sine & TIC & 6 & 0.124 & 0.104 & 2.471 & 0.086 & 0.951 & 0.536 & 1.719 & -0.025 & 0.219\\
8 & sine & TIC & 8 & 0.124 & 0.104 & 2.471 & 0.086 & 0.945 & 0.564 & 1.723 & -0.027 & 0.230\\
9 & sine & dcor2d\_2 & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.679\\
10 & sine & loess2d & 6 & 0.083 & 0.307 & 2.194 & 0.292 & 1.016 & 1.039 & 2.648 & 0.080 & 0.699\\
\addlinespace
11 & sine & splines2d & 6 & 0.040 & 0.189 & 2.606 & 0.104 & 1.014 & 1.051 & 2.730 & -0.009 & 0.727\\
12 & sine & stringy & 6 & 0.000 & 1173.035 & 1.031 & 17608.047 & 1.040 & 0.011 & 254.733 & 0.170 & 0.731\\
\bottomrule

@@

\begin{longtable}[t]{lrrrrrrl}

\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed
for modelling.}

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.394 & 0.207 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.394 & 0.207 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.394 & 0.207 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.394 & 0.207 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.394 & 0.207 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.394 & 0.100 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.394 & 0.100 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum numer of tries are
positively associated with the JS optimiser success rate while data
dimension and being flagged as long runtime are negatively associated
with teh success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.}

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.849 & 4.496 & -1.078 & 0.281\\
Smoothness & 1.672 & 1.527 & 1.094 & 0.274\\
Squintability & 6.967 & 2.173 & 3.207 & 0.001\\
dimension (d) & -0.588 & 0.257 & -2.292 & 0.022\\
long time & -0.816 & 1.333 & -0.612 & 0.540\\
\addlinespace
number of jellyfish & 0.229 & 0.131 & 1.752 & 0.080\\
maximum number of tries & 0.108 & 0.153 & 0.703 & 0.482\\
\bottomrule

\section{Additional plot:}\label{additional-plot}

@@
